---
title: "Introduction to bushfireRisk"
author: "Sravan Kumar Kapu Bejjala"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to bushfireRisk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `bushfireRisk` package provides tools and data for analyzing the attribution of Australian bushfire risk to anthropogenic climate change. This package is based on research by Oldenborgh et al. (2021), which demonstrated that human-caused climate change has significantly increased bushfire risk in Australia, particularly in the southeast region.

### What's Included

- **Clean datasets**: Ready-to-use data on Fire Danger Index, regional fire risk, and climate attributions
- **Analysis functions**: Calculate risk changes, plot trends, and compare regions
- **Interactive Shiny app**: Explore the data through an intuitive dashboard
- **Documentation**: Comprehensive guides and function references

## Installation

```{r eval=FALSE}
# Install from GitHub
remotes::install_github("ETC5523-2025/assignment-4-packages-Sravan9991")
```

## Quick Start

### Loading the Package

```{r eval=FALSE}
library(bushfireRisk)
library(dplyr)
library(ggplot2)
```

### Exploring the Data

The package includes three main datasets:

#### 1. bushfire_data

This is the primary dataset containing annual bushfire risk metrics for Australian regions from 1950 to 2020.

```{r eval=FALSE}
# Load the data
data("bushfire_data")

# View structure
str(bushfire_data)

# Summary statistics
summary(bushfire_data)

# View first few rows
head(bushfire_data)
```

The dataset includes:

- **year**: Year of observation (1950-2020)
- **region**: Australian region name
- **ffdi_mean**: Mean Forest Fire Danger Index
- **extreme_days**: Number of days with extreme fire danger
- **area_burned_1000ha**: Total area burned (thousands of hectares)
- **attribution_confidence**: Confidence that trends are due to climate change

#### 2. climate_attribution

This dataset compares fire danger under natural vs anthropogenic climate scenarios.

```{r eval=FALSE}
data("climate_attribution")
head(climate_attribution)
```

#### 3. regional_summary

Summary statistics by region across the full time period.

```{r eval=FALSE}
data("regional_summary")
regional_summary
```

## Basic Analysis

### Calculating Risk Change

Use `calculate_risk_change()` to find the percentage change in fire risk:

```{r eval=FALSE}
# FFDI change in Southeast Australia from 1950 to 2020
calculate_risk_change(
  data = bushfire_data,
  region = "Southeast Australia",
  start_year = 1950,
  end_year = 2020,
  metric = "ffdi_mean"
)

# Change in extreme fire days
calculate_risk_change(
  data = bushfire_data,
  region = "Southeast Australia",
  start_year = 1980,
  end_year = 2020,
  metric = "extreme_days"
)
```

### Plotting Trends

Create visualizations with `plot_ffdi_trend()`:

```{r eval=FALSE}
# Single region
plot_ffdi_trend(
  data = bushfire_data,
  regions = "Southeast Australia",
  add_trend = TRUE
)

# Compare multiple regions
plot_ffdi_trend(
  data = bushfire_data,
  regions = c("Southeast Australia", "Eastern Australia", "Southern Australia"),
  add_trend = TRUE
)

# Interactive plot
plot_ffdi_trend(
  data = bushfire_data,
  regions = "Southeast Australia",
  interactive = TRUE
)
```

### Regional Comparisons

Compare fire risk across regions:

```{r eval=FALSE}
# Recent comparison (2000-2020)
compare_regions(bushfire_data, year_range = c(2000, 2020))

# Historical comparison
compare_regions(bushfire_data, year_range = c(1950, 2020))
```

## Custom Analysis

### Example 1: Regional Fire Danger Over Time

```{r eval=FALSE}
library(ggplot2)
library(dplyr)

# Filter for Southeast Australia
southeast_data <- bushfire_data %>%
  filter(region == "Southeast Australia")

# Create plot
ggplot(southeast_data, aes(x = year, y = ffdi_mean)) +
  geom_line(color = "#d7191c", linewidth = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "#2c7bb6") +
  labs(
    title = "Fire Danger Trends in Southeast Australia",
    subtitle = "1950-2020",
    x = "Year",
    y = "Mean Forest Fire Danger Index",
    caption = "Data source: Oldenborgh et al. (2021)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "gray40")
  )
```

### Example 2: Extreme Fire Days by Region

```{r eval=FALSE}
# Calculate average extreme days by region (2000-2020)
extreme_summary <- bushfire_data %>%
  filter(year >= 2000) %>%
  group_by(region) %>%
  summarise(
    avg_extreme_days = mean(extreme_days, na.rm = TRUE),
    max_extreme_days = max(extreme_days, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_extreme_days))

# Bar plot
ggplot(extreme_summary, aes(x = reorder(region, avg_extreme_days), 
                            y = avg_extreme_days)) +
  geom_col(fill = "#e74c3c") +
  coord_flip() +
  labs(
    title = "Average Extreme Fire Days by Region",
    subtitle = "2000-2020 Average",
    x = NULL,
    y = "Average Number of Extreme Fire Days per Year"
  ) +
  theme_minimal()
```

### Example 3: Climate Attribution

```{r eval=FALSE}
# Plot natural vs anthropogenic scenarios
ggplot(climate_attribution, aes(x = year, y = ffdi_mean, 
                                color = scenario, linetype = scenario)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(
    values = c("Natural Climate" = "#2c7bb6", 
               "Anthropogenic Climate" = "#d7191c")
  ) +
  labs(
    title = "Climate Change Attribution Analysis",
    subtitle = "Natural vs Anthropogenic Climate Influence",
    x = "Year",
    y = "Forest Fire Danger Index",
    color = "Scenario",
    linetype = "Scenario"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Interactive Dashboard

Launch the Shiny app for interactive exploration:

```{r eval=FALSE}
launch_bushfire_app()
```

The dashboard includes:

- **Regional Analysis**: Compare fire risk across regions with interactive filtering
- **Time Trends**: Visualize long-term patterns with customizable time ranges
- **Attribution Analysis**: Explore the human influence on fire risk
- **Data Explorer**: Browse and download the complete dataset

## Key Findings

Based on the analysis, the package demonstrates:

1. **Southeast Australia shows the strongest increase** in fire risk, with FFDI increasing by approximately 40% from 1950 to 2020

2. **Human-caused climate change is responsible** for much of this increase, with 95% confidence in the attribution for Southeast Australia

3. **Extreme fire days are becoming more common**, with a clear upward trend especially since 1980

4. **Without emission reductions**, catastrophic fire seasons like the 2019-2020 "Black Summer" will become more frequent

## Next Steps

- Read the [Climate Attribution](climate-attribution.html) vignette for detailed methodology
- See [Using the App](using-the-app.html) for a complete dashboard guide
- Explore the function documentation: `?launch_bushfire_app`, `?plot_ffdi_trend`

## References

Oldenborgh, G.J., Philip, S.Y., Kew, S.F., van der Wiel, K., et al. (2021). Attribution of the Australian bushfire risk to anthropogenic climate change. *Environmental Research Letters*.
